Rates and delay times of type Ia supernovae in the Dark Energy Survey

### 2021-02-09 draft.tex
Updates from Overleaf

A complementary technique involves measuring the evolution of the volumetric rate of SNe Ia as a function of redshift and comparing this evolution to the average cosmic star-formation history (CSFH) of the Universe \citep{Gal-Yam2004,Strolger2004,Dahlen2004,Dahlen2008, Graur2018, Frohmaier2019}. This technique (the volumetric rate method) is also applicable to galaxy clusters, for which it is assumed that the star-formation histories are strongly peaked at some past epoch and thus that the rate of SNe Ia in clusters as a function of redshift is a more direct measure of the DTD \citep{Maoz2004,Graur2014,Maoz2014}. These studies have, almost ubiquitously, found $\beta$ to be consistent with $-1$.

@@

SNe Ia do not occur instantaneously after an episode of star formation. Instead stars with zero-age main-sequence (ZAMS) masses less than $8 \mathrm{M}_{\odot}$ take time to evolve along the main sequence and form WDs and this time is not well constrained. In many works this ``prompt time" $t_{\mathrm{p}}$ is fixed to some value expected to be the minimum time for a star to evolve off the MS and become a WD, such as 40~Myr \citep{Maoz2012,Graur2014,Graur2018}. Other studies have left $t_{\mathrm{p}}$ as a free parameter \citep{Heringer2019,Castrillo2020}. We perform two fits: one with $t_{\mathrm{p}}$ fixed at 40 Myr, and one with it as a free parameter. We fit $t_{\mathrm{p}}$ in log space and adopt a Gaussian prior with hyper-parameters $p(\log(t_{\mathrm{p}})) \sim \mathcal{N}(-1.3\,,0.2)$, which corresponds to being centred around 50~Myr. Choosing this regime to fit allows more prior weight to be placed on shorter prompt times as per the majority of the literature.

The posterior distributions for the fit parameters are found in Figs. \ref{fig:corner_beta_norm} and \ref{fig:corner_beta_norm_tp} for the fixed and free $t_{\mathrm{p}}$ fits respectively, and the posterior means and standard deviations are summarised in Table \ref{tab:dtd_results}. In Fig. \ref{fig:rate_fitted} we show the SN rate per galaxy predicted by the model assuming the best fitting DTD parameters from Table \ref{tab:dtd_results} compared to the data as presented in Section \ref{sec:rates}. The model provides a good fit across a wide range of stellar mass. The model clearly diverges from the simple log-linear fit, describing better the suppressed rate at the low and high ends of the stellar mass range, as well as the enhanced rate around $10^{10}$ M$_{\odot}$. This improvement provides evidence to support the DTD*SFH model. However, the model prediction still diverges from the data at low ($<10^9$ M$_{\odot}$) and high ($>10^{11.25}$ M$_{\odot}$), suggesting that these data cannot be explained by the our model, but are driven by further processes that we have not included. We discuss these in Section \ref{subsec:discussion}.

\textbf{Or's suggestion: Fix beta, A and fit for tp}

@@

\subsubsection{The DTD power law index \label{subsubsec:compare_beta}}
We measure a DTD power law index of $-0.96\pm0.05$. This value is consistent with the majority of previous analyses using various methods. Our measurement provides an independent support to the values close to $-1$ found using volumetric rates \citep[e.g.][]{Graur2018,Frohmaier2019}, individual galaxy SFHs \citep[e.g.][]{Maoz2012,Graur2013}, and galaxy clusters \citep[e.g.][]{Maoz2010} although our slope is marginally inconsistent with that of \citet{Heringer2019}.

\subsubsection{The DTD normalisation \label{subsubsec:compare_A}}
We measure a DTD normalisation (i.e. the rate of SNe 1 Gyr after star-formation) of $1.70 \pm 0.05 \times 10^{-13}$~M$_{\odot}^{-1}$~yr$^{-1}$. \citet{Graur2013} found a value of $7 \times 10^{-14}$~M$_{\odot}^{-1}$~yr$^{-1}$ using the SFH technique, whereas \citet{Heringer2019} report $7 \times 10^{-13}$~M$_{\odot}^{-1}$~yr$^{-1}$. These results lie either side of our measured value. \citet{Heringer2019} note that the normalisation recovered from the SFHR method is sensitive to the assumed prompt time $t_{\mathrm{p}}$ (see Section \ref{subsubsec:compare_tp}, which we also find using our method.

@@

The majority of works in the field have constrained the DTD power-law slope, and many also estimate its normalisation. Conversely, few have attempted to constrain the time after a burst of star-formation at which SNe Ia begin to explode -- the prompt time $t_{\mathrm{p}}$. In most cases, $t_{\mathrm{p}}$ has been fixed at some fiducial value such as 40~Myr, which is derived from the lifetime of 8~M$_{\odot}$ stars. \citet{Castrillo2020} included $t_{\mathrm{p}}$ (which they denote $\Delta$) in their fit, and find a value of $50_{-35}^{+100}$~Myr which is consistent with both 40~Myr as well as our best fitting value of $90 \pm 50$~Myr. \citet{Heringer2019} also present fits with varying prompt times, although they don't fit for the parameter itself. They find that varying $t_{\mathrm{p}}$ changes the recovered normalisation, which is consistent with the degeneracy seen in our results (Fig. \ref{fig:corner_beta_norm_tp}). We therefore report results that are consistent with the fiducial value of 40~Myr.

@@

The use of SN Ia rates to constrain the SN Ia DTD has consistently led to a $t^{-1}$ power-law, which is widely accepted as evidence for the DD scenario. While our results are also consistent with those derived in previous studies, our model is mildly inconsistent with the SN rate per galaxy at very low and very high stellar mass. In both cases, the model over predicts the observed rate of SNe Ia in galaxies in that stellar mass range, and thus any second order effects must act to suppress the rate compared to the fiducial model.

@@

Low mass galaxies are highly sensitive to environmental effects and may often be found in the so-called starburst state with abnormally high (above the SFMS) levels of star formation that could drive environmental evolution \citep[e.g.][]{Dekel1986,Marlowe1995,Tremonti2004,Lee2006}, although studies have shown that only 6 percent of dwarf galaxies are starbursts \citep{Lee2009}. It is possible that the averaging effects of the mass assembly model overestimate the true, instantaneous SFR for a significant fraction of the dwarf galaxies. Since the SN Ia rate is dominated by short delay times in these populations, an overestimated SFR in the model would lead to an overestimate in the SN Ia rate.

At some point along the evolutionary pathway, galaxies begin to cease star formation due to a combination of processes that together are known as quenching. In our model of mass assembly the characteristic mass at which quenching occurs is described by Eq. \ref{eq:mq}, and the rate of the transition from star forming to passive (as a function of the stellar mass) is determined by the transition scale $\sigma_{\mathrm{Q}}$ which we set to 1.1 based on GAMA observations. It is possible that this transition is too narrow and that quenching happens too fast in our model, leading to an under-prediction of the prompt fraction of SNe in the highest mass galaxies. To address this issue we rerun the mass assembly model with $\sigma_{\mathrm{Q}} = 1.5$ as per the nominal analysis of \citet{Childress2014} and refit the SN Ia rate data.

@@

The mass-metallicity relation (MZR) is a strong observed correlation between galaxy stellar mass and gas-phase metallicity \citep[e.g.][]{Tremonti2004}, whereby higher mass galaxies have undergone more cycles of stellar evolution and have been polluted with heavy elements created in stars and released in SNe. Thus, the mass of a host galaxy is inextricably linked to the metallicity of its hot gas. However, the nature of the DTD complicates matters, since stars of different ages were formed at different epochs where the galaxy had a different stellar mass and thus different metallicity. Therefore, there is not a direct correlation between the metallicity of SN Ia progenitors and their observed host galaxy stellar mass.  This caveat notwithstanding, it may be an expected consequence of the stipulations of e.g. \citet{Kistler2013} that the rate of SNe Ia is suppressed in the highest mass galaxies where the metallicity is highest, as is the case in our data. However, we would also expect to see this effect reversed at the low stellar mass end -- the rate should be enhanced in the lowest-mass, lowest-metallicity galaxies, yet instead we see a further suppression compared to the intermediate masses. Perhaps at low stellar mass, the effect of metallicity is negligible compared to the limitations of our SFH model.

@@

In the previous sections we have considered all SNe that passed the various quality, standardisation, and redshift cuts. However, it is well established that SNe Ia show correlations of varying strength between properties of their light curves and host galaxies. For example, measures of the light curve duration (e.g. decline rate $\Delta m_{15}$, or stretch such as SALT2 $x_1$) are known to correlate with host galaxy properties such as stellar mass \citep{Lampeitl2010,Sullivan2010}.

>>>>

A complementary technique involves measuring the evolution of the volumetric rate of SNe Ia as a function of redshift and comparing this evolution to the average cosmic star-formation history (CSFH) of the Universe \citep{Gal-Yam2004,Strolger2004,Dahlen2004,Dahlen2008, Graur2013, Frohmaier2019}. This technique (the volumetric rate method) is also applicable to galaxy clusters, for which it is assumed that the star-formation histories are strongly peaked at some past epoch and thus that the rate of SNe Ia in clusters as a function of redshift is a more direct measure of the DTD \citep{Maoz2004,Graur2014,Maoz2014}. These studies have, almost ubiquitously, found $\beta$ to be consistent with $-1$.

@@

SNe Ia do not occur instantaneously after an episode of star formation. Instead stars with zero-age main-sequence (ZAMS) masses less than $8~\mathrm{M}_{\odot}$ take time to evolve along the main sequence and form WDs and this time is not well constrained. In many works this ``prompt time" which we denote $t_{\mathrm{p}}$ is fixed to some value expected to be the minimum time for a star to evolve off the MS and become a WD, such as 40~Myr \citep{Maoz2012,Graur2013,Graur2014}. Other studies have left $t_{\mathrm{p}}$ as a free parameter \citep{Heringer2019,Castrillo2020}. We perform three fits: one with $t_{\mathrm{p}}$ fixed at 40 Myr, one with it as a free parameter along with $\beta$ and $A$, and a third with $\beta$ and $A$ fixed and  $t_{\mathrm{p}}$ free. We fit $t_{\mathrm{p}}$ in log space and adopt a Gaussian prior with hyper-parameters $p(\log(t_{\mathrm{p}})) \sim \mathcal{N}(-1.3\,,0.2)$, which corresponds to being centred around 50~Myr. Choosing this regime to fit allows more prior weight to be placed on shorter prompt times as per the majority of the literature.

The posterior distributions for the fit parameters are found in Figs. \ref{fig:corner_beta_norm} and \ref{fig:corner_beta_norm_tp} for the fixed and free $t_{\mathrm{p}}$ fits respectively, and the posterior means and standard deviations are summarised in Table \ref{tab:dtd_results}. In Fig. \ref{fig:rate_fitted} we show the SN rate per galaxy predicted by the model assuming the best fitting DTD parameters from Table \ref{tab:dtd_results} compared to the data as presented in Section \ref{sec:rates}. The model provides a good fit across a wide range of stellar mass. The model clearly diverges from the simple log-linear fit, describing better the suppressed rate at the low and high ends of the stellar mass range, as well as the enhanced rate around $10^{10}$ M$_{\odot}$. This improvement provides evidence to support the DTD*SFH model. However, the model prediction still diverges from the data at high ($>10^{11.25}$ M$_{\odot}$), suggesting that these data cannot be explained by the our model, but are driven by further processes that we have not included. We discuss these in Section \ref{subsec:discussion}.

@@

\subsubsection{The DTD power law index \label{subsubsec:compare_beta}}
We measure a DTD power law index of $-1.14\pm0.05$. This value is consistent with the majority of previous analyses using various methods. Our measurement provides an independent support to the values close to $-1$ found using volumetric rates \citep[e.g.][]{Graur2013,Frohmaier2019}, individual galaxy SFHs \citep[e.g.][]{Maoz2012,Graur2013}, and galaxy clusters \citep[e.g.][]{Maoz2010} although our slope is marginally inconsistent with that of \citet{Heringer2019}.

\subsubsection{The DTD normalisation \label{subsubsec:compare_A}}
We measure a DTD normalisation (i.e. the rate of SNe 1 Gyr after star-formation) of $1.73 \pm 0.05 \times 10^{-13}$~M$_{\odot}^{-1}$~yr$^{-1}$. \citet{Graur2013} found a value of $7 \times 10^{-14}$~M$_{\odot}^{-1}$~yr$^{-1}$ using the SFH technique, whereas \citet{Heringer2019} report $7 \times 10^{-13}$~M$_{\odot}^{-1}$~yr$^{-1}$. These results lie either side of our measured value. \citet{Heringer2019} note that the normalisation recovered from the SFHR method is sensitive to the assumed prompt time $t_{\mathrm{p}}$, a degeneracy that is also apparent our method.

@@

The majority of works in the field have constrained the DTD power-law slope, and many also estimate its normalisation. Conversely, few have attempted to constrain the time after a burst of star-formation at which SNe Ia begin to explode -- the prompt time $t_{\mathrm{p}}$. In most cases, $t_{\mathrm{p}}$ has been fixed at some fiducial value such as 40~Myr, which is derived from the lifetime of 8~M$_{\odot}$ stars. \citet{Castrillo2020} included $t_{\mathrm{p}}$ (which they denote $\Delta$) in their fit, and find a value of $50_{-35}^{+100}$~Myr which is consistent with both 40~Myr as well as our best fitting value of $67_{-23}^{+43}$~Myr for the three-parameter fit and $47_{-7}^{+8}$~Myr with $\beta$ and $A$ fixed. \citet{Heringer2019} also present fits with varying prompt times, although they don't fit for the parameter itself. They find that varying $t_{\mathrm{p}}$ changes the recovered normalisation, which is consistent with the degeneracy seen in our results (Fig. \ref{fig:corner_beta_norm_tp}). Overall, we find results for $t_{\mathrm{p}}$ that are consistent with the fiducial value of 40~Myr and adopt this value for the remainder of the analysis.

@@

The use of SN Ia rate measurements to constrain the SN Ia DTD has consistently led to a $t^{-1}$ power-law, which is widely accepted as evidence supporting the DD scenario. While our results are also consistent with those derived in previous studies, our model is mildly inconsistent with the SN rate per galaxy at very high stellar mass. The model over predicts the observed rate of SNe Ia in galaxies in that stellar mass range, caused either by a miscalculation of the rate or second order effects acting to suppress the rate compared to the fiducial model.

@@

At some point along the evolutionary pathway, galaxies begin to cease star formation due to a combination of processes that together are known as quenching. In our model of mass assembly the characteristic mass at which quenching occurs is described by Eq. \ref{eq:mq}, and the rate of the transition from star forming to passive (as a function of the stellar mass) is determined by the transition scale $\sigma_{\mathrm{Q}}$ which we set to 1.1 based on GAMA observations. It is possible that this transition is too narrow and that quenching happens too fast in our model, leading to an under-prediction of the prompt fraction of SNe in the highest mass galaxies. To address this issue we rerun the mass assembly model with $\sigma_{\mathrm{Q}} = 1.5$ as per the nominal analysis of \citet{Childress2014} and refit the SN Ia rate data. We find that the adapted quenching model does not change the ability of the model to adequately fit the high-mass turnover in the SN rate.

@@

The mass-metallicity relation (MZR) is a strong observed correlation between galaxy stellar mass and gas-phase metallicity \citep[e.g.][]{Tremonti2004}, whereby higher mass galaxies have undergone more cycles of stellar evolution and have been polluted with heavy elements created in stars and released in SNe. Thus, the mass of a host galaxy is inextricably linked to the metallicity of its hot gas. However, the nature of the DTD complicates matters, since stars of different ages were formed at different epochs where the galaxy had a different stellar mass and thus different metallicity. Therefore, there is not a direct correlation between the metallicity of SN Ia progenitors and their observed host galaxy stellar mass.  This caveat notwithstanding, it may be an expected consequence of the stipulations of e.g. \citet{Kistler2013} that the rate of SNe Ia is suppressed in the highest mass galaxies where the metallicity is highest, as is the case in our data.

@@

In the previous sections we have considered all SNe that passed the various quality, standardisation, and redshift cuts. However, it is well established that SNe Ia show correlations of varying strength between properties of their light curves and host galaxies. For example, measures of the light curve duration (e.g. decline rate $\Delta m_{15}$, or stretch such as SALT2 $x_1$) are known to correlate with host galaxy properties such as stellar mass \citep{Lampeitl2010,Sullivan2010} and specific SFR \citep{Rigault2018}. In this section we split the DES SNe Ia by their colour and stretch and assess how the measured rates and DTDs vary across this parameter space. To do so we split the sample into sub-samples with $x_1 = 0$ and $c=0$ as the divisions points respectively. We repeat the analyses of Sections \ref{sec:rates} and \ref{sec:model}, performing the DTD fitting with a fixed $t_{\mathrm{p}} = 40$~Myr, and present the results in Table \ref{tab:dtd_results_split_lc}.

\subsection{Splitting by SN stretch \label{subsec:split_x1}}

The SN Ia rate as a function of stellar mass and the best fitting model for SNe split by stretch is shown in Fig. \ref{fig:rate_fitted_split_x1}. The evolution of the rate with stellar mass is significantly different for the two sub-samples: high-stretch SNe dominate the rate at low stellar mass but tail off in the higher mass galaxies, while low-stretch SNe are subdominant in low mass galaxies but display a much stronger correlation with mass and make up nearly all SNe in high-mass hosts.
